---
title: 'Installing Hortonworks Sanbox Deployment (HDP) on Docker Mac'
author: 'Konrad'
date: '2019-02-23'
slug: installing-hortonworks-sanbox-deployment-hdp-on-docker-mac
categories:
  - how-to
  - setup
tags:
  - docker
  - bash
---

# Background

The post covers installation of Hortonworks Sandbox Deployment (HDP) on Mac using Docker. 

## Installation

### Docker

Before installing docker let's check for the existing installation.

```{r docker_check, echo=TRUE, eval=FALSE}
which -a docker
# /usr/local/bin/docker
```

Assuming that the line above did not produce results we can install docker with use of homebrew. To install homebrew:

```{r brew_install, eval=FALSE, echo=TRUE}
/usr/bin/ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)"
```

we can the progress with installing docker

```{r docker_install, eval=FALSE, echo=TRUE}
brew install docker
```

### HDP

The HDP can be installed with the provided scrip. The script can be download from 



